---
title: "ESG Company SQL Exploratory Data Analysis"
author: "Josie Zenger"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This report is an exploratory data analysis (EDA) of ESG (Environmental, Social, and Governance) assessment data for a variety of companies. The analysis uses SQL queries to extract insights from two databases downloaded from Kaggle.com that contain company profiles, ESG risk ratings, controversy scores, and other company-specific metrics. The purpose of this report is to showcase my SQL skills in querying, analyzing, and visualizing data from multiple tables to derive meaningful insights.

Skills exhibited: Data cleaning, joins, aggregations, filtering, and visualization (plotly and ggplot).

# Data Overview

The dataset comprises two main tables from Kaggle.com: the [S&P 500 ESG Risk Ratings]("https://www.kaggle.com/datasets/pritish509/s-and-p-500-esg-risk-ratings")" dataset compiled by Pritish Dugar in 2023, and [Public Company ESG Ratings Dataset]("https://www.kaggle.com/datasets/alistairking/public-company-esg-ratings-dataset") dataset compiled by Alistair King in 2023. These datasets are publically available. The datasets contain the following information:

(1). **`Public Company ESG Ratings Dataset`**: 
  (a). *Basic company information*: ticker symbol, company name, currency, exchange, industry, logo URL, website URL
  (b). *Environmental scores and rating*: environment_score, environment_grade, environment_level
  (c). *Social scores and rating*: social_score, social_grade, social_level
  (d). *Governance scores and rating*: governance_score, governance_grade, governance_level
  (e). *Overall ESG scores and rating*: total_score, total_grade, total_level
  (f). Last processing date of the ESG data
  (g). CIK identifier

(2). **`S&P 500 ESG Risk Ratings`**: Symbol, Name, Address, Sector, Industry, Full Time Employees, Description, Total ESG Risk Score, Environment Risk Score, Governance Risk Score, Social Risk Score, Controversy Level, Controversy Score, ESG Risk Percentile, ESG Risk Level

```{r Load Libraries and Data, echo=FALSE}
# Load required libraries
library(dplyr) 
library(sqldf)  
library(readr)  
library(DBI)
library(tidyverse)
library(plotly)
library(DT)

# Load the data files
public_esg <- read.csv("~/Local/Personal_Projects/SQL/data.csv") 
sp_esg <- read.csv("~/Local/Personal_Projects/SQL/SP_Risk_Ratings.csv")  

```

```{r Clean Data, echo=FALSE}
# Rename columns: Replace periods with underscores and convert to lowercase
sp_esg <- sp_esg %>%
  rename_with(~ gsub("\\.", "_", .x) %>% tolower())
```

# SQL Queries and Visualizations

### Evaluate Similarities Between the Datasets
The first query evaluates the similarities between the two datasets by joining them on the company ticker symbol and comparing the industry fields. The query uses the **'LIKE'** function to compare companies with similar industry labels in both datasets.
```{r Evaluate Dataset, echo=TRUE}
# Define the SQL query to join the two tables
query0 <- "
SELECT d1.name AS Name, d1.industry AS D1_Industry, d2.industry AS D2_Industry
FROM public_esg d1
INNER JOIN sp_esg d2
ON d2.symbol = UPPER(d1.ticker)
WHERE 
    d1.industry LIKE '%' || d2.industry || '%'
    OR d2.industry LIKE '%' || d1.industry || '%'
ORDER BY d1.industry
"

# Execute the query
eval_dataset <- sqldf(query0)

# Display the result in a datatable
datatable(eval_dataset, options = list(
  scrollX = TRUE,  
  pageLength = 5   
))
```

### ESG Risk Scores by Company
The following query retrieves the ESG risk scores (total, environmental, social, and governance) for companies, sorted alphabetically by name. The results are displayed in an interactive plot using  **plotly**.
```{r Query 1, echo=TRUE}
query1<- 
"SELECT 
    d1.name AS company_name, d1.industry, total_esg_risk_score, environment_risk_score, governance_risk_score, social_risk_score
FROM 
    public_esg d1
INNER JOIN 
    sp_esg d2
ON 
    d2.symbol = UPPER(d1.ticker)
WHERE total_esg_risk_score IS NOT NULL
ORDER BY d1.name ASC"

# Execute the query
esg_risk_scores <- sqldf(query1)
# Display the result in a datatable
datatable(esg_risk_scores, options = list(
  scrollX = TRUE,  
  pageLength = 5   
))

```

```{r Plot Query 1, echo=FALSE, fig.width=9, fig.height=6, fig.align='center'}

# Plot the ESG risk scores by company
plot1 <- plot_ly(
  data = esg_risk_scores, # Drawing data from Query 1
  x = ~`total_esg_risk_score`, 
  y = ~`company_name`, 
  type = 'scatter',
  mode = 'markers',
  text = ~paste("Company:", company_name, "<br>Total ESG Risk Score:", total_esg_risk_score), # Tooltip text
  hoverinfo = 'text', # Only shows company and ESG risk rating in the tooltip
  marker = list(size = 10, color = 'black') # Default color for all points
) %>%
  layout(
    title = list(
      text = "Interactive Plot of Total ESG Risk Scores by Company",
      font = list(size = 20) # Increase title size
    ),
    # X-axis label formatting
    xaxis = list(
      title = "Total ESG Risk Score",
      titlefont = list(size = 18),
      ticktext = c("Low", "Medium", "High"), # Optional tick labels
      showticklabels = TRUE
    ),
    # Y-axis label formatting
    yaxis = list(
      title = "Company",
      showticklabels = FALSE,
      titlefont = list(size = 18)
    ),
    # Add annotation for citation
    annotations = list(
      list(
        text = "Source: Kaggle.com", # Citation
        x = 0.1, y = -0.1,  # Position of the annotation
        showarrow = FALSE,
        font = list(size = 10),
        xref = "paper",
        yref = "paper",
        xanchor = "left",
        yanchor = "top"
      )
    ),
    # Adjust margins for the plot
    margin = list(
      l = 100, # Left margin
      r = 50,  # Right margin
      t = 100, # Top margin to avoid title cutoff
      b = 100  # Bottom margin for citation
    )
  )

# Display the plot
plot1

```


### Average ESG Risk by Industry
The following query calculates the average ESG risk score for industries with a low ESG risk level. The results are displayed in a bar plot using **ggplot**. I use the **'ROUND'** and **'GROUP BY'** functions to calculate the average risk score for each industry.
```{r Query 2, echo=TRUE}
query2 <- 
"SELECT 
    d1.industry AS Industry, ROUND(avg(total_esg_risk_score), 2) AS Average_Industry_Risk
FROM 
    public_esg d1
INNER JOIN 
    sp_esg d2
ON 
    d2.symbol = UPPER(d1.ticker)
WHERE esg_risk_level = 'Low'
GROUP BY d1.industry
ORDER BY Average_Industry_Risk ASC"

# Execute the query
avg_industry <- sqldf(query2)
```

```{r Plot Query 2, echo=FALSE, fig.width=9, fig.height=6, fig.align='center'}

# Reorder the bars in descending order
avg_industry <- avg_industry %>%
  mutate(Industry = reorder(Industry, -Average_Industry_Risk))

# Create the bar plot
ggplot(avg_industry, aes(x = Industry, y = Average_Industry_Risk)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Average ESG Risk by Industry",
    x = "",
    y = "Average ESG Risk Score" ,
    caption = "Source: Kaggle.com" 
  ) +
  theme_minimal() +
  coord_flip()  # Flip the coordinates for better readability
```
**Findings**: The plot shows the average ESG risk scores for industries with a low ESG risk level. The 'Auto Components' industry has the lowest average ESG risk score, while the 'Machinery' industry has the highest average ESG risk score.

### Companies with High Controversy Scores
The following query identifies companies with controversy scores exceeding 3, sorted by controversy level. The results are displayed in a datatable using the **DT** package.
```{r Query 3, echo=TRUE}
query3 <-
"SELECT 
    d1.name as Name, controversy_level AS Controversy_Level, controversy_score AS Controversy_Score, d1.industry as Industry
FROM 
    public_esg d1
INNER JOIN 
    sp_esg d2
ON 
    d2.symbol = UPPER(d1.ticker)
WHERE controversy_score > 3
ORDER BY controversy_level"

# Execute the query
high_contro <- sqldf(query3)

datatable(high_contro, options = list(
  scrollX = TRUE,  
  pageLength = 5   
))
```
**Findings**: Companies with high controversy scores come from a variety of industries, including automobiles, media, and banking. The ESG practices of these companies are controversial and should not be replicated in an ideal ESG framework.

### Companies with Exceptional Governance Scores
The following query identifies companies with governance scores exceeding 300 (from the public_esg dataset) and governance risk scores (from the sp_esg dataset) below 5. The results are displayed in a datatable using the **DT** package.
```{r Query 4, echo=TRUE}
query4 <-
"SELECT 
    d1.name As Name, governance_score As Governance_Score, governance_risk_score As Governance_Risk_Score
FROM 
    public_esg d1
INNER JOIN 
    sp_esg d2
ON 
    d2.symbol = UPPER(d1.ticker)
WHERE governance_score > 300 
    AND governance_risk_score < 5
ORDER BY governance_score DESC"

# Execute the query
high_govern <- sqldf(query4)

# Format the output using the datatable function
datatable(high_govern, options = list(
  scrollX = TRUE,  # horizontal scrolling
  pageLength = 5   # Set the number of rows per page
))

```
**Findings:** There is some discrepency between the governance scores and governance risk scores. Companies with high governance scores may not necessarily have low governance risk scores. This could be due to differences in the scoring methodologies used by the two datasets. However, these companies have effective governance practices that should be emulated by others.

### Processing Dates for Large Companies with Over 75,000 Employees
```{r Query 5, echo=TRUE}
# Remove commas from 'full_time_employees' and convert it to numeric
sp_esg$full_time_employees <- as.numeric(gsub(",", "", sp_esg$full_time_employees))

# Convert 'last_processing_date' to Date format 
public_esg$last_processing_date <- as.Date(public_esg$last_processing_date, format = "%d-%m-%Y")

# Run the query
query5 <- "
SELECT 
    d1.name AS Name, last_processing_date AS Processing_Date, full_time_employees AS Full_Time_Employees
FROM 
    public_esg d1
INNER JOIN 
    sp_esg d2
ON 
    d2.symbol = UPPER(d1.ticker)
WHERE full_time_employees > 75000
  AND last_processing_date = '19098' OR last_processing_date = '19099'
ORDER BY last_processing_date DESC
"

# Execute the query
large_companies <- sqldf(query5)

# Convert last-processing-date back into date format
large_companies$Processing_Date <- as.Date(large_companies$Processing_Date, origin = "1970-01-01")

# Format the output using the datatable function
datatable(large_companies, options = list(
  scrollX = TRUE,  # horizontal scrolling
  pageLength = 5   # Set the number of rows per page
))

```

### Companies in High ESG Risk Percentiles
The following query categorizes companies into ESG risk percentiles (Top 10%, Top 25%, Top 50%, Bottom 50%) and ranks them. The results are displayed in a datatable using the **DT** package and an interactive plot using the **plotly** package.
```{r Query 6, echo=TRUE}
query6 <-
"SELECT *
FROM (
    SELECT 
        name AS Name, 
        esg_risk_percentile AS ESG_Risk_Percentile,
        CASE
            WHEN CAST(SUBSTRING(esg_risk_percentile, 1, 2) AS INT) >= 90 THEN 'Top 10%'
            WHEN CAST(SUBSTRING(esg_risk_percentile, 1, 2) AS INT) BETWEEN 75 AND 89 THEN 'Top 25%'
            WHEN CAST(SUBSTRING(esg_risk_percentile, 1, 2) AS INT) BETWEEN 50 AND 74 THEN 'Top 50%'
            WHEN CAST(SUBSTRING(esg_risk_percentile, 1, 2) AS INT) < 50 THEN 'Bottom 50%'
        END AS Category
    FROM 
        sp_esg
) subquery
WHERE Category IS NOT NULL
    AND Category = 'Top 10%' OR Category = 'Top 25%' or Category = 'Top 50%'
ORDER BY esg_risk_percentile DESC"

# Execute the query
esg_percentile <- sqldf(query6)

# Format the output using the datatable function
datatable(esg_percentile, options = list(
  scrollX = TRUE,  # horizontal scrolling
  pageLength = 5   # Set the number of rows per page
))

```

```{r Plot Query 3, echo=FALSE, fig.width=9, fig.height=6, fig.align='center'}

# Define the percentiles
top_10_percentile <- 36
top_25_percentile <- 23
top_50_percentile <- 0

# Plot the ESG risk scores by company
plot2 <- plot_ly(
  data = esg_percentile, # Drawing data from Query 1
  x = ~`ESG_Risk_Percentile`, 
  y = ~`Name`, 
  type = 'scatter',
  mode = 'markers',
  text = ~paste("Company:", Name, "<br>ESG Risk Percentile:", ESG_Risk_Percentile), 
  hoverinfo = 'text', 
  marker = list(size = 10, color = 'black') 
) %>%
  layout(
    title = list(
      text = "Interactive Plot of Companies in High ESG Risk Percentiles",
      font = list(size = 20) # Increase title size
    ),
    # X-axis label formatting
    xaxis = list(
      title = "ESG Risk Percentile",
      titlefont = list(size = 18),
       # Optional tick labels
      showticklabels = FALSE
    ),
    # Y-axis label formatting
    yaxis = list(
      title = "Company",
      showticklabels = FALSE,
      titlefont = list(size = 18)
    ),
    # Add vertical lines for top percentiles
    shapes = list(
      list(
        type = "line",
        x0 = top_50_percentile,
        x1 = top_50_percentile,
        y0 = 0,
        y1 = 1,
        xref = "x", 
        yref = "paper",
        line = list(color = "green", width = 2, dash = "dash")
      ),
      list(
        type = "line",
        x0 = top_25_percentile,
        x1 = top_25_percentile,
        y0 = 0,
        y1 = 1,
        xref = "x", 
        yref = "paper",
        line = list(color = "orange", width = 2, dash = "dash")
      ),
      list(
        type = "line",
        x0 = top_10_percentile,
        x1 = top_10_percentile,
        y0 = 0,
        y1 = 1,
        xref = "x", 
        yref = "paper",
        line = list(color = "skyblue", width = 2, dash = "dash")
      )
    ),
    # Add labels for the vertical lines
    annotations = list(
      # Label for Top 50%
      list(
        text = "Top 50%", # Label text
        x = top_50_percentile, # X-position of the label
        y = 1.01,  # Y-position of the label (just above the plot)
        showarrow = TRUE,
        arrowhead = 2,
        ax = 0,
        ay = -30,
        font = list(size = 14, color = "green"),
        xref = "x", 
        yref = "paper"
      ),
      # Label for Top 25%
      list(
        text = "Top 25%", # Label text
        x = top_25_percentile, # X-position of the label
        y = 1.01,  # Y-position of the label
        showarrow = TRUE,
        arrowhead = 2,
        ax = 0,
        ay = -30,
        font = list(size = 14, color = "orange"),
        xref = "x", 
        yref = "paper"
      ),
      # Label for Top 10%
      list(
        text = "Top 10%", # Label text
        x = top_10_percentile, # X-position of the label
        y = 1.01,  # Y-position of the label
        showarrow = TRUE,
        arrowhead = 2,
        ax = 0,
        ay = -30,
        font = list(size = 14, color = "skyblue"),
        xref = "x", 
        yref = "paper"
      ),
      # Citation
      list(
        text = "Source: Kaggle.com", # Citation
        x = 0.1, y = -0.2,  # Position of the citation
        showarrow = FALSE,
        font = list(size = 10),
        xref = "paper",
        yref = "paper",
        xanchor = "left",
        yanchor = "top"
      )
    ),
    updatemenus = list(
        list(
          buttons = list(
            list(method = "restyle", args = list("marker.color", 
                list(ifelse(esg_percentile$Category == "Top 10%", 'skyblue', 'black'))), # Filter Global Top 50
              label = "Top 10%"),
            list(method = "restyle", args = list("marker.color", 
                list(ifelse(esg_percentile$Category == "Top 25%", 'orange', 'black'))), # Filter Industry Top Rated
              label = "Top 25%"),
            list(method = "restyle", args = list("marker.color", 
                list(ifelse(esg_percentile$Category == "Top 50%", 'lightgreen', 'black'))), # Filter Region Top Rated
              label = "Top 50%"),
            list(method = "restyle",args = list("marker.color", list('black')), # Reset to original color
              label = "Show All")),
          direction = "up", x = 0.1, y = -0.1, showactive = TRUE, type = "dropdown")),
    # Adjust margins for the plot
    margin = list(
      l = 100, # Left margin
      r = 50,  # Right margin
      t = 100, # Top margin to avoid title cutoff
      b = 100  # Bottom margin for citation
    )
  )

# Display the plot
plot2

```
**Findings:** Companies in the Top 10% have the highest ESG risk scores, while those in the top 50% have lower ESG risk scores. The Top 10% category has overlap with the list of companies created with high controversy scores. This plot can assist companies to set benchmarks for their ESG practices based on their intended ESG Risk Percentiles. 


# Conclusion

This analysis is an exploratory data analysis of ESG Risk data from two different datasets with data visualizations. My SQL queries extract various insights, such as similarities between the datasets, ESG risk scores by company, average ESG risk by industry, companies with high controversy scores, companies with exceptional governance scores, processing dates for large companies, and companies in high ESG risk percentiles. The visualizations provide an interactive and informative way to explore the data and derive meaningful insights.

Future steps include:

1. Expanding the analysis to include time-series trends.
2. Comparing ESG scores with financial performance metrics.
3. Developing predictive models for ESG risk assessment.
